\bf Manifold Clustering in the Generalized Random Dot Product Graph

### 2022-04-27 draft/draft.Rmd
still working ...

\begin{definition}[Manifold block model]
Let $p, q \geq 0$, $d = p + q \geq 1$, $K \geq 1$, and $n \geq 1$ be integers.
If there are manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \subset \mathcal{X}$ for $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top I_{p,q} y \in [0, 1] \}$, and 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ such that each $x_i \in \mathcal{M}_k$ for some $k \in [K]$, then $A \sim \GRDPG_{p,q}(X; \rho_n)$ is a \emph{manifold block model}.
\end{definition} 
In practice, we define each manifold $\mathcal{M}_k$ by a continuous function $g_k : [0, 1]^r \to \mathcal{X}$, where $1 \leq r < d$ is the dimensionality of manifold $\mathcal{M}_k$.

Insert something about ASE and its consistency here.

\begin{example}
Let $p = 2$, $q = 0$, $K = 2$, and $r = 1$. 
Define two manifolds by $f_1(t) = \begin{bmatrix} \cos(\frac{\pi}{3} t) & \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$ and $f_2(t) = \begin{bmatrix} 1 - \cos(\frac{\pi}{3} t) & 1 - \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$.
Draw $t_1, ..., t_n \sim \Uniform(0, 1)$ and $z_1, ..., z_n \iid \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. 
Finally, let $A \sim \RDPG(X)$.

```{r fig1, fig.cap = 'Manifold block model described in Example 1. The latent configuration is on the left, and a random dot product graph drawn from the latent configuration is on the middle, and the ASE is on the right.', fig.height = 2.5, fig.width = 7}
set.seed(123456)
n1 <- 2 ** 5
n2 <- 2 ** 5
n <- n1 + n2
z <- c(rep(1, n1), rep(2, n2))
u1 <- runif(n1)
u2 <- runif(n2)
x1 <- cos(pi / 3 * u1)
y1 <- sin(pi / 3 * u1)
x2 <- 1 + cos(pi / 3 * u2 + pi)
y2 <- 1 + sin(pi / 3 * u2 + pi)
data.matrix <- cbind(c(x1, x2), c(y1, y2))
A <- draw.graph(data.matrix %*% t(data.matrix))
Xhat <- embedding(A, 2, 0)

@@

## Nonintersecting Manifolds

In this section, we consider the following scenario: Suppose that each community is represented by a manifold $\mathcal{M}_k$, $k \in \{1, ..., K\}$ in the latent space of a RDPG or GDRPG, and the manifolds do not intersect each other. 
Define $\delta = \min\limits_{k, \ell} \min\limits_{x \in \mathcal{M}_k, y \in \mathcal{M}_\ell} \|x - y\|$, the minimum distance between two manifolds. 
We assume that $\delta > 0$.

First, we consider the case in which each $\mathcal{M}_k \subset \mathbb{R}^d$ is a one-dimensional manifold, which can be characterized by a function $g_k : [0, 1] \to \mathcal{M}_k$. 
Let $F$ be a probability distribution with support $[0, 1]$. 
Then we define a mixture model as follows:

1. Draw $t_1, ..., t_n \iid F$.
2. Draw $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$, the community labels.
3. Let each $x_i = g_{z_i}(t_i)$ be the latent vector for vertex $v_i$.  
Assume $\min_{k, \ell} \min_{t, s} \|g_k(t) - g_\ell(s)\| = \delta > 0$. 
4. Draw $A \sim \RDPG(X)$ or $A \sim \GRDPG_{p,q}(X)$.

\begin{theorem}[Community detection for one-dimensional nonintersecting manifolds without noise]
\label{nonintersect-no-noise}
Let $x_1, ..., x_n$ be points sampled on $K$ one-dimensional manifolds $\mathcal{M}_1, ..., \mathcal{M}_K$, defined by arc-length parameterizations $g_1, ..., g_K$. 
Suppose $\delta = \min_{k, \ell} \min_{t, s} \| g_k(t) - g_\ell(s) \| > 0$. 
Let $A_n$ be the event such that $\max_k \max_i \|x_{(i+1)}^{(k)} - x_{(i)}^{(k)}\| > \delta$, where $x_{(i)}^{(k)}$ is the $i^{th}$ order statistic on the $k^{th}$ manifold. 
Then for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, $P(A_n) < \epsilon$. 
\end{theorem}

\begin{theorem}[Community detection for one-dimensional nonintersecting manifolds with noise]
\label{nonintersect-with-noise}
Suppose the same setup as in theorem \ref{nonintersect-no-noise}. 
Define $y_i = x_i + e_i$ and $B_n$ as the event such that $\|y_{(i+1)}^{(k)} - y_{(i)}^{(k)}\| > \|y_{(i)}^{(k)} - y_{(j)}^{(\ell)}\|$ for some $i$, $j$, $k$, and $\ell$. 
Then if $\max_i \|e_i\| < \delta / 3$, for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, $P(B_n) < \epsilon$. 
\end{theorem}

\begin{theorem}[Community detection for one-dimensional nonintersecting manifold block models]
\label{nonintersect-mbm}
Suppose we have the same setup as in theorem \ref{nonintersect-no-noise}. 
Let $p, q \geq 0$ such that $p + q = d$ where $d$ is the dimension of the range of each $g_k$. 
Suppose we draw $A \sim \GRDPG_{p,q}(X; \rho_n)$ for some sparsity parameter $\rho_n \in (0, 1]$, and let $\hat{X}$ be the ASE of $A$. 
Then for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, single linkage clustering produces zero community detection error with probability 1. 
\end{theorem}

@@

## Intersecting Manifolds

In this section, we again consider the setting for the RDPG or GRDPG in which each community lies on a manifold in the latent space. 
However, this time, we do not assume that the manifolds are nonintersecting.

@@

\begin{proof}[Proof of theorem \ref{nonintersect-no-noise}]
This theorem requires two additional conditions: 
First, each $n_k \to \infty$ as $n \to \infty$, and second, the number of manifolds $K$ remains fixed. 
These are satisified by the model assumptions. 
Then this is a direct consequence of Lemma \ref{lem:lim-d}. 
Denoting $m = \min_k n_k$ as the sample size of the manifold with the fewest points,

$$
\begin{aligned}
P(\max_{k,i} \|X_{(i+1)}^{(k)} - X_{(i)}^{(k)}\| < \delta) & = P(\forall k, i ~ D_i^{(k)} < \delta) \\
& \geq \prod_{k=1}^K P(D_i^{(k)} < \delta) \\
& \geq (1 - (m-1) (1 - a \delta)^{m} + (m-1) (1 - F(1-\delta) - a \delta)^{m})^K
\end{aligned}
$$
Since as $n \to \infty$, $m \to \infty$ as well, and $K$ is fixed, this approaches $(1)^K = 1$. 
\end{proof}

\begin{proof}[Proof of theorem \ref{nonintersect-with-noise}]
\end{proof}

>>>>

\begin{definition}[Manifold block model]
Let $p, q \geq 0$, $d = p + q \geq 1$, $K \geq 1$, and $n \geq 1$ be integers.
If there are manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \subset \mathcal{X}$ for $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top I_{p,q} y \in [0, 1] \}$, and 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ such that each $x_i \in \mathcal{M}_k$ for some $k \in [K]$, then $A \sim \GRDPG_{p,q}(X; \rho_n)$ is a \emph{manifold block model}.
\end{definition}

In practice, we often define each manifold $\mathcal{M}_k$ by a continuous function $g_k : [0, 1]^r \to \mathcal{X}$, where $1 \leq r < d$ is the dimensionality of manifold $\mathcal{M}_k$. 
To complete the probability model, we define a probability distribution $F$ with support $[0, 1]^r$ and draw the community memberships from a multinomial distribution.

The full mixture model can be described as follows:

1. Draw labels $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$. 
2. Draw latent vectors by first drawing each $t_1, ..., t_n \iid F$ and then computing each $x_i = g_{z_i}(t_i)$. 
3. Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$, 
and draw $A \sim \RDPG(X)$ or $A \sim \GRDPG_{p,q}(X)$.

\begin{example}
Let $p = 2$, $q = 0$, $K = 2$, and $r = 1$. 
Define two one-dimensional manifolds by $f_1(t) = \begin{bmatrix} \cos(\frac{\pi}{3} t) & \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$ and $f_2(t) = \begin{bmatrix} 1 - \cos(\frac{\pi}{3} t) & 1 - \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$.
Draw $t_1, ..., t_n \iid \Uniform(0, 1)$ and $z_1, ..., z_n \iid \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. 
Finally, let $A \sim \RDPG(X)$. Fig \ref{fig:example1} shows the latent configuration drawn from this latent distribution, a random dot product graph drawn from the latent configuration, and the adjacency spectral embedding of the graph.

```{r example1, fig.cap = 'Manifold block model described in Example 1. The latent configuration is on the left, and a random dot product graph drawn from the latent configuration is on the middle, and the ASE is on the right.', fig.height = 2.5, fig.width = 7}
set.seed(123456)
n1 <- 2 ** 5
n2 <- 2 ** 5
n <- n1 + n2
z <- c(rep(1, n1), rep(2, n2))
u1 <- runif(n1)
u2 <- runif(n2)
x1 <- cos(pi / 3 * u1)
y1 <- sin(pi / 3 * u1)
x2 <- 1 + cos(pi / 3 * u2 + pi)
y2 <- 1 + sin(pi / 3 * u2 + pi)
data.matrix <- cbind(c(x1, x2), c(y1, y2))
A <- draw.graph(data.matrix %*% t(data.matrix))
Xhat <- embedding(A, 2, 0)

@@

## Nonintersecting Manifolds
\label{section:nonintersecting}

In this section, we consider the following scenario: 
Suppose that each community is represented by a manifold $\mathcal{M}_k$, $k \in \{1, ..., K\}$ in the latent space of a RDPG or GDRPG, and the manifolds do not intersect each other. 
Define $\delta = \min\limits_{k \neq \ell} \min\limits_{x \in \mathcal{M}_k, y \in \mathcal{M}_\ell} \|x - y\|$, the minimum distance between two manifolds. 
We assume that $\delta > 0$, i.e., the manifolds do not intersect.

In the noiseless setting, if the subsample on each manifold is sufficiently dense, it is possible to construct for each manifold an $\eta_k$-neighborhood graph for each manifold for some $\eta_k > 0$ such that the graph is connected. 
Then if $\max_k \eta_k = \eta < \delta$, an $\eta$-neighborhood graph for the entire sample will consist of $K$ disconnected subgraphs that map onto each manifold. 
Equivalently, we can apply single-linkage clustering. 
The remainder of this section explores under which conditions these criteria are met for the latent configuration, in which latent vectors lie exactly on manifolds, as well as the ASE, which introduces noise.

@@

Consider the specific case in which each $\mathcal{M}_k \subset \mathbb{R}^d$ is a one-dimensional manifold. 
Let $F_k$ be a probability distribution with support $\mathcal{M}_k$. 
Then we define a mixture model as follows:

1. Draw labels $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$. 
2. Draw latent vectors each as $x_i \indep F_{z_i}$ for distributions $F_1, ..., F_K$ with respective supports $\mathcal{M}_1, ..., \mathcal{M}_K$. 
3. Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$, 
and draw $A \sim \RDPG(X)$ or $A \sim \GRDPG_{p,q}(X)$.

Note that here, we redefine the model to ignore $g_1, ..., g_K$, the parameterizations of each manifold. 
Instead, we sample points directly on the manifolds themselves. 
We will return to the parameterizations in Section \ref{section:intersecting}.

In this setting, although each manifold is in $d$-dimensional space, since the manifolds are one-dimensional, it is possible to consider an ordering of points sampled on the manifold. 
In the original latent space in which there is no noise, the condition for which community detection is possible via $\eta$-neighborhood graph clustering or single linkage clustering is each pair of adjacent order statistics on each manifold being less than $\delta$ apart.

\begin{theorem}[Community detection for one-dimensional nonintersecting manifolds without noise]
\label{nonintersect-no-noise}
Let $x_1, ..., x_n$ be points sampled on $K$ one-dimensional manifolds $\mathcal{M}_1, ..., \mathcal{M}_K$. 
Suppose $\delta = \min_{k \neq \ell} \min_{x_i \in \mathcal{M}_k, x_j \in \mathcal{M}_\ell} \| x_i - x_j \| > 0$. 
Let $A_n$ be the event such that $\max_k \max_i \|x_{(i+1)}^{(k)} - x_{(i)}^{(k)}\| < \delta$, where $x_{(i)}^{(k)}$ is the $i^{th}$ order statistic on the $k^{th}$ manifold. 
Then for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, $P(A_n) > 1 - \epsilon$. 
\end{theorem}

Sampling a graph from this latent configuration and constructing an ASE from the graph introduces error relative to the original latent vectors. 
If the errors are small, which is guaranteed by the consistency property of the ASE for sufficiently large $n$, then  sufficient separation between embedding points belonging to different manifolds is still guaranteed.

\begin{theorem}[Community detection for one-dimensional nonintersecting manifolds with noise]
\label{nonintersect-with-noise}
Suppose the same setup as in theorem \ref{nonintersect-no-noise}. 
Define $y_i = x_i + e_i$ and $B_n$ as the event such that $\|y_{(i+1)}^{(k)} - y_{(i)}^{(k)}\| < \|y_{(i)}^{(k)} - y_{(j)}^{(\ell)}\|$ for any $k \neq \ell$. 
Here, $y_{(i)}^{(k)}$ is the $i^{th}$ order statistic on the $k^{th}$ manifold plus its corresponding noise. 
Then if $\max_i \|e_i\| < \delta / 3$, for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, $P(B_n) > 1 - \epsilon$. 
\end{theorem}

\begin{theorem}[Community detection for one-dimensional nonintersecting manifold block models]
\label{nonintersect-mbm}
Suppose we have the same setup as in theorem \ref{nonintersect-no-noise}. 
Let $p, q \geq 0$ such that $p + q = d$ where $d$ is the dimension of the range of each $g_k$. 
Suppose we draw $A \sim \GRDPG_{p,q}(X; \rho_n)$ for some sparsity parameter $\rho_n \in (0, 1]$, and let $\hat{X}$ be the ASE of $A$. 
Then for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, single linkage clustering produces zero community detection error with high probability. 
\end{theorem}

## Intersecting Manifolds
\label{section:intersecting}

In this section, we again consider the setting for the RDPG or GRDPG in which each community lies on a manifold in the latent space. 
However, this time, we do not assume that the manifolds are nonintersecting. 
We also restrict this case to one-dimensional manifolds which are each described by $g_k : [0, 1] \to \mathcal{X}$. 
Then we define a mixture model as follows:

1. Draw $t_1, ..., t_n \iid F$ for probability distribution $F$ with support $[0, 1]$.
2. Draw $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$, the community labels.
3. Let each $x_i = g_{z_i}(t_i)$ be the latent vector for vertex $v_i$, and collect the latent vectors into matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
4. Draw $A \sim \RDPG(X)$ or $A \sim \GRDPG_{p,q}(X)$.

@@

\begin{proof}[Proof of theorem \ref{nonintersect-no-noise}]
This theorem requires two additional conditions: 
First, each $n_k \to \infty$ as $n \to \infty$, and second, the number of manifolds $K$ remains fixed. 
These are satisfied by the model assumptions. 
Then this is a direct consequence of Lemma \ref{lem:lim-d}. 
Denoting $m = \min_k n_k$ as the sample size of the manifold with the fewest points,

$$
\begin{aligned}
P(\max_{k,i} \|X_{(i+1)}^{(k)} - X_{(i)}^{(k)}\| < \delta) & = P(\forall k, i ~ D_i^{(k)} < \delta) \\
& \geq \prod_{k=1}^K P(D_i^{(k)} < \delta) \\
& \geq (1 - (m-1) (1 - a \delta)^{m} + (m-1) (1 - F(1-\delta) - a \delta)^{m})^K
\end{aligned}
$$
Since as $n \to \infty$, $m \to \infty$ as well, and $K$ is fixed, this approaches $(1)^K = 1$. Then setting this $\geq 1 - \epsilon$, it is always possible to solve this for $m$. 
\end{proof}

\begin{proof}[Proof of theorem \ref{nonintersect-with-noise}]
Let $\max_i \|e_i\| \leq \xi$. 
\end{proof}

### 2022-04-27 draft/draft.tex
still working ...

In practice, we define each manifold \(\mathcal{M}_k\) by a continuous
function \(g_k : [0, 1]^r \to \mathcal{X}\), where \(1 \leq r < d\) is
the dimensionality of manifold \(\mathcal{M}_k\).

Insert something about ASE and its consistency here.

\begin{example}
Let $p = 2$, $q = 0$, $K = 2$, and $r = 1$. 
Define two manifolds by $f_1(t) = \begin{bmatrix} \cos(\frac{\pi}{3} t) & \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$ and $f_2(t) = \begin{bmatrix} 1 - \cos(\frac{\pi}{3} t) & 1 - \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$.
Draw $t_1, ..., t_n \sim \mathrm{Uniform}(0, 1)$ and $z_1, ..., z_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Multinomial}(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. 
Finally, let $A \sim \mathrm{RDPG}(X)$.

@@

{\centering \includegraphics{draft_files/figure-latex/fig1-1}

@@

\caption{Manifold block model described in Example 1. The latent configuration is on the left, and a random dot product graph drawn from the latent configuration is on the middle, and the ASE is on the right.}\label{fig:fig1}
\end{figure}
\end{example}

@@

In this section, we consider the following scenario: Suppose that each
community is represented by a manifold \(\mathcal{M}_k\),
\(k \in \{1, ..., K\}\) in the latent space of a RDPG or GDRPG, and the
manifolds do not intersect each other. Define
\(\delta = \min\limits_{k, \ell} \min\limits_{x \in \mathcal{M}_k, y \in \mathcal{M}_\ell} \|x - y\|\),
the minimum distance between two manifolds. We assume that
\(\delta > 0\).

First, we consider the case in which each
\(\mathcal{M}_k \subset \mathbb{R}^d\) is a one-dimensional manifold,
which can be characterized by a function
\(g_k : [0, 1] \to \mathcal{M}_k\). Let \(F\) be a probability
distribution with support \([0, 1]\). Then we define a mixture model as
follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Draw \(t_1, ..., t_n \stackrel{\mathrm{iid}}{\sim}F\).
\item
  Draw
  \(z_1, ..., z_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Multinomial}(\alpha_1, ..., \alpha_K)\),
  the community labels.
\item
  Let each \(x_i = g_{z_i}(t_i)\) be the latent vector for vertex
  \(v_i\).\\
  Assume
  \(\min_{k, \ell} \min_{t, s} \|g_k(t) - g_\ell(s)\| = \delta > 0\).
\item
  Draw \(A \sim \mathrm{RDPG}(X)\) or
  \(A \sim \mathrm{GRDPG}_{p,q}(X)\).
\end{enumerate}

\begin{theorem}[Community detection for one-dimensional nonintersecting manifolds without noise]
\label{nonintersect-no-noise}
Let $x_1, ..., x_n$ be points sampled on $K$ one-dimensional manifolds $\mathcal{M}_1, ..., \mathcal{M}_K$, defined by arc-length parameterizations $g_1, ..., g_K$. 
Suppose $\delta = \min_{k, \ell} \min_{t, s} \| g_k(t) - g_\ell(s) \| > 0$. 
Let $A_n$ be the event such that $\max_k \max_i \|x_{(i+1)}^{(k)} - x_{(i)}^{(k)}\| > \delta$, where $x_{(i)}^{(k)}$ is the $i^{th}$ order statistic on the $k^{th}$ manifold. 
Then for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, $P(A_n) < \epsilon$. 
\end{theorem}

\begin{theorem}[Community detection for one-dimensional nonintersecting manifolds with noise]
\label{nonintersect-with-noise}
Suppose the same setup as in theorem \ref{nonintersect-no-noise}. 
Define $y_i = x_i + e_i$ and $B_n$ as the event such that $\|y_{(i+1)}^{(k)} - y_{(i)}^{(k)}\| > \|y_{(i)}^{(k)} - y_{(j)}^{(\ell)}\|$ for some $i$, $j$, $k$, and $\ell$. 
Then if $\max_i \|e_i\| < \delta / 3$, for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, $P(B_n) < \epsilon$. 
\end{theorem}

\begin{theorem}[Community detection for one-dimensional nonintersecting manifold block models]
\label{nonintersect-mbm}
Suppose we have the same setup as in theorem \ref{nonintersect-no-noise}. 
Let $p, q \geq 0$ such that $p + q = d$ where $d$ is the dimension of the range of each $g_k$. 
Suppose we draw $A \sim \mathrm{GRDPG}_{p,q}(X; \rho_n)$ for some sparsity parameter $\rho_n \in (0, 1]$, and let $\hat{X}$ be the ASE of $A$. 
Then for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, single linkage clustering produces zero community detection error with probability 1. 
\end{theorem}

@@

@@

In this section, we again consider the setting for the RDPG or GRDPG in
which each community lies on a manifold in the latent space. However,
this time, we do not assume that the manifolds are nonintersecting.

>>>>

In practice, we often define each manifold \(\mathcal{M}_k\) by a
continuous function \(g_k : [0, 1]^r \to \mathcal{X}\), where
\(1 \leq r < d\) is the dimensionality of manifold \(\mathcal{M}_k\). To
complete the probability model, we define a probability distribution
\(F\) with support \([0, 1]^r\) and draw the community memberships from
a multinomial distribution.

The full mixture model can be described as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Draw labels
  \(z_1, ..., z_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Multinomial}(\alpha_1, ..., \alpha_K)\).
\item
  Draw latent vectors by first drawing each
  \(t_1, ..., t_n \stackrel{\mathrm{iid}}{\sim}F\) and then computing
  each \(x_i = g_{z_i}(t_i)\).
\item
  Let \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\), and
  draw \(A \sim \mathrm{RDPG}(X)\) or
  \(A \sim \mathrm{GRDPG}_{p,q}(X)\).
\end{enumerate}

\begin{example}
Let $p = 2$, $q = 0$, $K = 2$, and $r = 1$. 
Define two one-dimensional manifolds by $f_1(t) = \begin{bmatrix} \cos(\frac{\pi}{3} t) & \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$ and $f_2(t) = \begin{bmatrix} 1 - \cos(\frac{\pi}{3} t) & 1 - \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$.
Draw $t_1, ..., t_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Uniform}(0, 1)$ and $z_1, ..., z_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Multinomial}(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. 
Finally, let $A \sim \mathrm{RDPG}(X)$. Fig \ref{fig:example1} shows the latent configuration drawn from this latent distribution, a random dot product graph drawn from the latent configuration, and the adjacency spectral embedding of the graph.

@@

{\centering \includegraphics{draft_files/figure-latex/example1-1}

@@

\caption{Manifold block model described in Example 1. The latent configuration is on the left, and a random dot product graph drawn from the latent configuration is on the middle, and the ASE is on the right.}\label{fig:example1}
\end{figure}
\end{example}

@@

\label{section:nonintersecting}

In this section, we consider the following scenario: Suppose that each
community is represented by a manifold \(\mathcal{M}_k\),
\(k \in \{1, ..., K\}\) in the latent space of a RDPG or GDRPG, and the
manifolds do not intersect each other. Define
\(\delta = \min\limits_{k \neq \ell} \min\limits_{x \in \mathcal{M}_k, y \in \mathcal{M}_\ell} \|x - y\|\),
the minimum distance between two manifolds. We assume that
\(\delta > 0\), i.e., the manifolds do not intersect.

In the noiseless setting, if the subsample on each manifold is
sufficiently dense, it is possible to construct for each manifold an
\(\eta_k\)-neighborhood graph for each manifold for some \(\eta_k > 0\)
such that the graph is connected. Then if
\(\max_k \eta_k = \eta < \delta\), an \(\eta\)-neighborhood graph for
the entire sample will consist of \(K\) disconnected subgraphs that map
onto each manifold. Equivalently, we can apply single-linkage
clustering. The remainder of this section explores under which
conditions these criteria are met for the latent configuration, in which
latent vectors lie exactly on manifolds, as well as the ASE, which
introduces noise.

@@

Consider the specific case in which each
\(\mathcal{M}_k \subset \mathbb{R}^d\) is a one-dimensional manifold.
Let \(F_k\) be a probability distribution with support
\(\mathcal{M}_k\). Then we define a mixture model as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Draw labels
  \(z_1, ..., z_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Multinomial}(\alpha_1, ..., \alpha_K)\).
\item
  Draw latent vectors each as
  \(x_i \stackrel{\mathrm{ind}}{\sim}F_{z_i}\) for distributions
  \(F_1, ..., F_K\) with respective supports
  \(\mathcal{M}_1, ..., \mathcal{M}_K\).
\item
  Let \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\), and
  draw \(A \sim \mathrm{RDPG}(X)\) or
  \(A \sim \mathrm{GRDPG}_{p,q}(X)\).
\end{enumerate}

Note that here, we redefine the model to ignore \(g_1, ..., g_K\), the
parameterizations of each manifold. Instead, we sample points directly
on the manifolds themselves. We will return to the parameterizations in
Section \ref{section:intersecting}.

In this setting, although each manifold is in \(d\)-dimensional space,
since the manifolds are one-dimensional, it is possible to consider an
ordering of points sampled on the manifold. In the original latent space
in which there is no noise, the condition for which community detection
is possible via \(\eta\)-neighborhood graph clustering or single linkage
clustering is each pair of adjacent order statistics on each manifold
being less than \(\delta\) apart.

\begin{theorem}[Community detection for one-dimensional nonintersecting manifolds without noise]
\label{nonintersect-no-noise}
Let $x_1, ..., x_n$ be points sampled on $K$ one-dimensional manifolds $\mathcal{M}_1, ..., \mathcal{M}_K$. 
Suppose $\delta = \min_{k \neq \ell} \min_{x_i \in \mathcal{M}_k, x_j \in \mathcal{M}_\ell} \| x_i - x_j \| > 0$. 
Let $A_n$ be the event such that $\max_k \max_i \|x_{(i+1)}^{(k)} - x_{(i)}^{(k)}\| < \delta$, where $x_{(i)}^{(k)}$ is the $i^{th}$ order statistic on the $k^{th}$ manifold. 
Then for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, $P(A_n) > 1 - \epsilon$. 
\end{theorem}

Sampling a graph from this latent configuration and constructing an ASE
from the graph introduces error relative to the original latent vectors.
If the errors are small, which is guaranteed by the consistency property
of the ASE for sufficiently large \(n\), then sufficient separation
between embedding points belonging to different manifolds is still
guaranteed.

\begin{theorem}[Community detection for one-dimensional nonintersecting manifolds with noise]
\label{nonintersect-with-noise}
Suppose the same setup as in theorem \ref{nonintersect-no-noise}. 
Define $y_i = x_i + e_i$ and $B_n$ as the event such that $\|y_{(i+1)}^{(k)} - y_{(i)}^{(k)}\| < \|y_{(i)}^{(k)} - y_{(j)}^{(\ell)}\|$ for any $k \neq \ell$. 
Here, $y_{(i)}^{(k)}$ is the $i^{th}$ order statistic on the $k^{th}$ manifold plus its corresponding noise. 
Then if $\max_i \|e_i\| < \delta / 3$, for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, $P(B_n) > 1 - \epsilon$. 
\end{theorem}

\begin{theorem}[Community detection for one-dimensional nonintersecting manifold block models]
\label{nonintersect-mbm}
Suppose we have the same setup as in theorem \ref{nonintersect-no-noise}. 
Let $p, q \geq 0$ such that $p + q = d$ where $d$ is the dimension of the range of each $g_k$. 
Suppose we draw $A \sim \mathrm{GRDPG}_{p,q}(X; \rho_n)$ for some sparsity parameter $\rho_n \in (0, 1]$, and let $\hat{X}$ be the ASE of $A$. 
Then for any $\epsilon > 0$, there exists an $N \in \mathbb{N}$ such that when $n > N$, single linkage clustering produces zero community detection error with high probability. 
\end{theorem}

@@

\label{section:intersecting}

In this section, we again consider the setting for the RDPG or GRDPG in
which each community lies on a manifold in the latent space. However,
this time, we do not assume that the manifolds are nonintersecting. We
also restrict this case to one-dimensional manifolds which are each
described by \(g_k : [0, 1] \to \mathcal{X}\). Then we define a mixture
model as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Draw \(t_1, ..., t_n \stackrel{\mathrm{iid}}{\sim}F\) for probability
  distribution \(F\) with support \([0, 1]\).
\item
  Draw
  \(z_1, ..., z_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Multinomial}(\alpha_1, ..., \alpha_K)\),
  the community labels.
\item
  Let each \(x_i = g_{z_i}(t_i)\) be the latent vector for vertex
  \(v_i\), and collect the latent vectors into matrix
  \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\).
\item
  Draw \(A \sim \mathrm{RDPG}(X)\) or
  \(A \sim \mathrm{GRDPG}_{p,q}(X)\).
\end{enumerate}
